Exploring mutation-induced protein fold switching using multisequence Monte Carlo simulations

### 2017-05-12 Manuscript.tex
Minor changes

\section{Methods}
\subsection{Coarse-grained 3-letter model for protein folding}
\noindent
All calculations were carried out on a coarse-grained model for protein folding~\citenum{Bhattacherjee2012}. In this model, there are 3 different amino acid types: hydrophobic (h), polar (p) and turn-type (t). The backbone chain is represented atomistically by the N, H, $\mathrm{C}_\alpha$, $\mathrm{H}_{\alpha 1}$, C$'$ and O atoms. By contrast, the sidechain represention is simplified to a single enlarged $\mathrm{C}_\beta$ atom, which is geometrically identical for h and p types. The sidechain is absent for the t type which instead has an $\mathrm{H}_{\alpha 2}$ atom. The t type is therefore closely related to glycine. All bond lengths, bond angles, and peptide plane angles (180$^\circ$) are held fixed. Hence, an $N$-amino acid chain conformation $\rbar$ can, for any sequence $\sbar$, is therefore described by the set of 2$N$ backbone torsional angles $\{\phi_i$, $\psi_i\}_{i=1}^{N}$.

@@

\subsection{Sequence sets}
\noindent

\begin{table}
\caption{\label{tab:table1} Model sequences of different sequence lengths $N$ studied in this work. In addition to the 6 sequences listed, we studied sets of 1,024 sequences derived from the A1-N1 and A2-TN pairs, respectively, (see text) denoted S16 and S35, respectively, and a restricted S16-set of 144 sequences denoted $\mathrm{S16}_{144}$ (also studied in Ref.~\protect\citep{Holzgrafe2014}).  }
\begin{ruledtabular}
\begin{tabular}{lcr}
Name & $N$ & Sequence\\
\hline
N1 & 16 & phphphpttphphphp \\
A1 & 16 & pphpphhpphpphhpp \\
R1 & 16 & \textbf{(add)}\\
R2 & 16 &  \textbf{(add)}\\
A2 & 35 & (A1)ttt(A1)\\
TN & 35 & (A1)ttt(N1)\\
\end{tabular}
\end{ruledtabular}
\end{table}

\subsection{Monte Carlo simulation parameters}
\noindent
Both ST and MS simulations are carried out with two types of conformational updates $\rbar\rightarrow\rbar'$: (1) a global pivot move (20\%) which randomly picks a $\phi_\mathrm{i}$ angle or $\psi_\mathrm{i}$ angle and assigned a new value between $-\pi$ and $\pi$; and (2) an approximately local move (80\%) which turns the $\phi_\mathrm{i}$ and $\psi_\mathrm{i}$-angles of 4 consecutive amino acids in a coordinated manner.~\cite{Favrin2003} In MS simulations, sequence updates $\sbar\rightarrow\sbar'$ are attempted every $1000$ step and carried out by randomly picking a new sequence $\sbar'\ne\sbar$ and applying the accept-reject criterion in Eq.~\ref{accrej}. In ST, temperature updates $m\rightarrow m'$ are attempted every 100 steps. Observables are sampled every 10 or 100 MC step. All simulations carried out in this work are summarized in Table~1.

\begin{table}
\caption{\label{tab:table1} List of simulations carried out in this work. }
\begin{ruledtabular}
\begin{tabular}{lccccr}
Runs & Algorithm & $\kb T$  & MC steps &  Sequences\\
\hline
32 & ST & 0.43--0.65 & $8\times 10^6$ &A1\\ 
32 & ST & 0.43--0.65 & $8\times 10^6$ &N1\\ 
32 & ST & 0.43--0.65 & $8\times 10^6$ &R1\\ 
32 & ST & 0.43--0.65 & $8\times 10^6$ &R2\\ 
32$\times$8\footnote{32  runs per temperature for 8 temperatures in the range given.} & MS &0.43--0.65& $144\times 10^6$ & $\mathrm{S16}_{144}$\\
16 & MS & 0.43  & $5\times 10^9$ &  S16 \\
16 & MS & 0.46 & $4\times 10^9$ &  S35 \\
\end{tabular}
\end{ruledtabular}
\end{table}

\subsection{Stability measures}
\noindent
To quantify the stabilities of various folds, we use the measures used in Ref.~\cite{Holzgrafe2015}. We classify conformations $r$ as part of a fold if $Q\ge0.8$ where $Q\in [0,1]$ quantifies the similarity of $r$ to that fold. For folds A and B, $Q$ is determined by defining a set of backbone-backbone hydrogen bonds that are present in those folds (ignoring N and C terminal amino acids). For fold A, the fold-specific set of contacts (i,j) between the CO group of amino acid i and the NH group of amino acid j is $\{$(2,6), (3,7), (4,8), (5,9), (6,10), (7,11), (8,12), (9,13), (10,14), (11,15)$\}$, and for fold B the contact set is $\{$(3,14), (5,12), (7,10), (10,7), (12,5), (14,3)$\}$. A hydrogen bond is considered formed if (i) the NHO and COH angles both are $>90^\circ$ and (ii) the HO distance is $<2.7$~{\AA}.

@@

\begin{figure}
\includegraphics[width=6.0cm]{MCtraj}
\caption{Example of a multisequence simulation.}
\end{figure}

@@

\subsection{Stability properties of fold-to-fold mutational pathways}
\noindent 
An apparently general characteristic of designed and natural protein folding switching is that the stability of sequences near the switch point have reduced stability.~\cite{Bryan2010} The results from our model exhibit a similar trend. Fig.~6A and B shows the total stability $\Ptot$ at different mutational distances between start and end points. Intermediate sequences are less stable than at distances $h=0$ and $h=10$ although there are large variations as indicated by the spread between upper and lower bounds. Nonetheless, there is a clear statistical trend that sequences become less stability as more and more mutations are applied to the any of the 4 starting points. The minimum in stability occur closer to N1 or TN indicating that the all-$\alpha$ folds IA and IIA are more mutationally robust than the $\beta$-structure containing folds IB and IIB.

@@

The smooth average stability trends in Fig.~6A and B belie the real character of the individual mutational pathways which abruptly switch from one fold to another. The switch in fold along individual pathways are typically completed in only 1-2 mutations. This can be seen from the distribution of switch lengths, $L_\mathrm{s}$, in Fig.~5C and D, taken over all pathways with $\Pcut=0.50$, where  $L_\mathrm{s} = 2$ for a switch that goes directly from a node $P_A>P_B$ to a node $P_B>P_A$ and $L_\mathrm{s} > 2$ if it also passes through one or more bistable sequences. Hence, for $\Pcut = 0.5$, it is fairly common that viable pathways pass through one or more sequences are that bistable, i.e., that have substantial populations of both folds.

>>>>

\section{Model and Methods}
\subsection{Coarse-grained 3-letter model for protein folding}
\noindent
All calculations were carried out using the coarse-grained model for protein folding developed in Ref.~\citenum{Bhattacherjee2012}. In this model, there are 3 different amino acid types: hydrophobic (h), polar (p) and turn-type (t). The backbone chain is represented atomistically by the N, H, $\mathrm{C}_\alpha$, $\mathrm{H}_{\alpha 1}$, C$'$ and O atoms. By contrast, the sidechain represention is simplified to a single enlarged $\mathrm{C}_\beta$ atom, which is geometrically identical for h and p types. The sidechain is absent for the t type which instead has an $\mathrm{H}_{\alpha 2}$ atom. The t type is therefore closely related to glycine. All bond lengths, bond angles, and peptide plane angles (180$^\circ$) are held fixed. Hence, an $N$-amino acid chain conformation $\rbar$ can, for any sequence $\sbar$, is therefore described by the set of 2$N$ backbone torsional angles $\{\phi_i$, $\psi_i\}_{i=1}^{N}$.

@@

\subsection{Model sequences}

The sequences studied in this work are summarized in Table~I.

\begin{table}
\caption{\label{tab1} List of model sequences of different lengths $N$ studied in this work.}
\begin{ruledtabular}
\begin{tabular}{lcr}
Name & $N$ & Sequence\footnote{In addition, we study two sets of 1,024 sequences derived from the A1-N1 and A2-TN pairs (see text) denoted $\mathrm{S16}_{1024}$  and $\mathrm{S35}_{1024}$, respectively, and a restricted set $\mathrm{S16}_{144}$ with 144 sequences taken from S16 ($\mathrm{S16}_{144}$ was also studied in Ref.~\protect\citenum{Holzgrafe2014}). } \\
\hline
A1 & 16 & pphpphhpphpphhpp \\
N1 & 16 & phphphpttphphphp \\
R1 & 16 & \textbf{(add)}\\
R2 & 16 &  \textbf{(add)}\\
A2 & 35 & (A1)ttt(A1)\\
TN & 35 & (A1)ttt(N1)\\
\end{tabular}
\end{ruledtabular}
\end{table}

\subsection{Monte Carlo simulation parameters}
\noindent
Both ST and MS simulations are carried out with two types of conformational updates $\rbar\rightarrow\rbar'$: (1) a global pivot move (20\%) which randomly picks a $\phi_\mathrm{i}$ angle or $\psi_\mathrm{i}$ angle and assigned a new value between $-\pi$ and $\pi$; and (2) a semi-local move (80\%) which turns the $\phi_\mathrm{i}$ and $\psi_\mathrm{i}$-angles of 4 consecutive amino acids in a coordinated manner.~\cite{Favrin2001} In MS simulations, sequence updates $\sbar\rightarrow\sbar'$ are carried out by randomly picking a new sequence $\sbar'\ne\sbar$ and applying the accept-reject criterion in Eq.~\ref{accrej}. A sequence move is attempted every $1,000$ MC step while temperature updates $m\rightarrow m'$ are attempted every 100 steps. The simulations carried out in this work are summarized in Tables~II.

\begin{table}
\caption{\label{tab2} List of simulations carried out in this work. }
\begin{ruledtabular}
\begin{tabular}{lccccr}
Runs & Algorithm & $\kb T$  & MC steps &  Sequences\\
\hline
32 & ST & 0.43--0.65 & $8\times 10^6$ &A1\\ 
32 & ST & 0.43--0.65 & $8\times 10^6$ &N1\\ 
32 & ST & 0.43--0.65 & $8\times 10^6$ &R1\\ 
32 & ST & 0.43--0.65 & $8\times 10^6$ &R2\\ 
32$\times$8\footnote{32  runs per temperature for 8 temperatures in the range given.} & MS &0.43--0.65& $144\times 10^6$ & $\mathrm{S16}_{144}$\\
16 & MS & 0.43  & $5\times 10^9$ &  $\mathrm{S16}_{1024}$ \\
16 & MS & 0.46 & $4\times 10^9$ &  $\mathrm{S35}_{1024}$ \\
\end{tabular}
\end{ruledtabular}
\end{table}

\subsection{Stability measures}
\noindent
To quantify the stabilities of various folds, we use the measures introduced in Ref.~\cite{Holzgrafe2015}. We classify conformations $r$ as part of a fold if $Q\ge0.8$ where $Q\in [0,1]$ quantifies the similarity of $r$ to that fold.

For folds A and B, $Q$ is determined by defining a set of backbone-backbone hydrogen bonds that are present in those folds (ignoring N and C terminal amino acids). For fold A, the fold-specific set of contacts (i,j) between the CO group of amino acid i and the NH group of amino acid j is $\{$(2,6), (3,7), (4,8), (5,9), (6,10), (7,11), (8,12), (9,13), (10,14), (11,15)$\}$, and for fold B the contact set is $\{$(3,14), (5,12), (7,10), (10,7), (12,5), (14,3)$\}$. A hydrogen bond is considered formed if (i) the NHO and COH angles both are $>90^\circ$ and (ii) the HO distance is $<2.7$~{\AA}.

@@

\begin{figure}
\includegraphics[width=6.0cm]{MCtraj}
\caption{Example of a multisequence simulation trajectory.}
\end{figure}

@@

\subsection{Stability properties of fold-to-fold mutational pathways}
\noindent 
An apparently general characteristic of designed and natural protein folding switching is that the stability of sequences near the switch point have reduced stability.~\cite{Bryan2010} The results from our model exhibit a similar trend. Fig.~6A and B shows the total stability $\Ptot$ at different mutational distances between start and end points. Intermediate sequences are less stable than at distances $h=0$ and $h=10$ although there are large variations as indicated by the spread between upper and lower bounds. Nonetheless, there is a clear statistical trend that sequences become less stable as successive mutations are applied to any of the 4 starting points until a minimum is reached. These minima in stability occur closer to N1 and TN in the two systems, respectively, indicating that the all-$\alpha$ folds IA and IIA are more mutationally robust than the all-$\beta$ fold IB and mixed $\alpha$/$\beta$-fold IIB.

@@

The smooth average stability trends in Fig.~6A and B, however, belie the real character of the \textit{individual} mutational pathways which tend to exhibit abrupt switches between the two folds. 
%To examine how the fold transitions take place in our model, we classify all stable sequences (i.e. those with $\Ptot>\Pcut$) as either having one dominant fold, i.e., one of the two folds have little stability, or as bistable meaning that both folds have roughly similar stability $\PaI$ and $\PaII$.
This can be seen from the distributions of switch lengths, $L_\mathrm{s}$, in Fig.~5C and D, which are taken over all pathways with $\Pcut=0.50$,
%In determining $L_\mathrm{s}$, we define for each pathway how many nodes are included in the  ...
where  $L_\mathrm{s} = 2$ for a switch that goes directly from a node $P_A>P_B$ to a node $P_B>P_A$ and $L_\mathrm{s} > 2$ if it also passes through one or more bistable sequences. The switch in fold along individual pathways are typically completed in only 1-2 mutations. Hence, for $\Pcut = 0.5$, it is fairly common that viable pathways pass through one or more sequences are that bistable, i.e., that have substantial populations of both folds.
